Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change

### 2024-02-06 ms.qmd
Fixed Shin's comments and a few more of Franks. Checked with Shinichi about comparison of acclimation and relative variance.

Climate change is expected to result in warmer and more variable thermal environments globally. Greater thermal variability is expected to result in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity. Such responses depend on genetic and phenotypic variability, yet most work has focused on plastic changes in mean phenotypic responses to climate warming ignore how temperature may change phenotypic variability. Here, using a large database of physiological rate data from marine, freshwater and terrestrial ectotherms we quantify the capacity for temperature acclimation and the extent to which relative variance (changes in variance after accounting for changes in the mean) in physiological rates changes in response to increasing temperatures. Using new effect size estimates and meta-analysis we show that acclimation of physiological rates and changes in their relative variance do not depend on past temperature variability and predictability experienced by a population at its geographical location; neither do these responses differ between life stages. Across habitats relative variance in physiological rates decreased at higher temperatures. Freshwater ectotherms are capable of acclimating which is likely to be beneficial considering that reductions in relative variance in physiological rates at high temperatures is lower for freshwater species. In contrast, acclimation responses of marine organisms coincide with a nearly equal reduction in relative variance in physiolgical rates, suggesting acclimation may not necssarily be beneficial. Relative variance reductions are particularly pronounced for terrestrial ectotherms and this coincides with a lack of capacity for acclimation, which highlights the vulnerability of terrestrial ectotherms to climate change. Our results emphasise how beneficial acclimation responses may be offset by reductions in physiological rate variance, challenging the notion that plasticity will be sufficient to counteract climate change effects on populations. Reductions in physiological rate variance at high temperatures may further have important ecological ramifications that impact population and ecosystem processes that cascade to affect population and community productivity.

@@

Reversible phenotypic plasticity, such as acclimation, is expected to provide greater potential to buffer populations from climate impacts as responses are relatively rapid and can therefore be fine-tuned to proximate environmental conditions [assuming the costs of plasticity are low – @Dewitt1998; @Scheiner2020]. Acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to faciliate a rapid response to the environment [@seebacher2019histone; @little2013thyroid; @taff2016endocrine]. However, the focus up to now has been primarily on mean physiological responses. For example, mean thermal tolerances or acclimation abilities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022]. However, it is possible that intrapopulation variability might also be impacted in addition to the mean. Understanding how variability in physiological rates -- traits thought to be closely linked to fitness -- are affected by climate change is important because lack of physiological variation can limit responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variance in a population may also indicate greater niche breadth which can buffer populations against environmental change [@Bolnick2011; @Schindler2010; @Zheng2023; @Slatyer2013]. Decreases in phenotypic variance also suggests strong stabilising selection or reflect constraints on performance [@Scheiner2020; @Dewitt1998]. Changes in physiological trait variation may also have important ecological consequences by promoting population productivity and stability [@Agashe2009; @kooijman1989population], species coexistence and ecosystem processes [@imura2003genetic; @Bolnick2011; @hart2016variation]. The implications of changes in variance could, therefore, have wide-reaching consequences for understanding the capacity of populations to persist in and adapt to novel environments but to date have been largely ignored [@Scheiner2020].

@@

Nonetheless, effect heterogeneity was high (only `r mean(tot_i2)`% of the variance was the result of sampling variability, 95% CI: `r quantile(tot_i2, 0.025)[1]` to `r quantile(tot_i2, 0.975)[1]`%), and most variance was explained by the specific study and type of trait (Study: `r mean(stdy_i2)`% , 95% CI: `r quantile(stdy_i2, 0.025)[1]` to `r quantile(stdy_i2, 0.975)[1]`%; Trait Type: `r mean(trait_i2)`% , 95% CI: `r quantile(trait_i2, 0.025)[1]` to `r quantile(trait_i2, 0.975)[1]`%). Evolutionary relationships among taxa and species ecology (i.e., species random effect) explained little variation in acute and acclimation repsonses (Species: `r mean(spp_i2)`% , 95% CI: `r quantile(spp_i2, 0.025)[1]` to `r quantile(spp_i2, 0.975)[1]`%; Phylogeny: `r mean(phylo_i2)`% , 95% CI: `r quantile(phylo_i2, 0.025)[1]` to `r quantile(phylo_i2, 0.975)[1]`%).

@@

```{r, fig-1}
#| label: fig-1
#| fig-cap: "Taxonomic distribution of acute and acclimation $Q_{10}$ estimates across major habitats. **A)** Phylogenetic distribution of taxa contained within the data. The total number of acute and acclimation type Q10 effect sizes are highlighted as well as whether the taxa is marine, freshwater or terrestrial. Silhouettes are representative taxa of major clades within the tree. **B)** Acute and Acclimation $lnRR_{Q_{10}}$ across marine, freshwater, and terrestrial systems. **C)** $lnCVR_{Q_{10}}$ across traits for marine, freshwater and terrestrial systems. Note there were no differences between acute and acclimation $Q_{10}$ types. k = total number of effect size estimates while the numbers in brackets indicate the number of species."
#| warning: false
#| echo: false

@@

The total proportion of heterogeneity in $lnCVR_{Q_{10}}$ was lower compared to $lnRR_{Q_{10}}$ ($I^2_{sv}$ = `r mean(tot_i2_cvr)`, 95% CI: `r quantile(tot_i2_cvr, 0.025)` to `r quantile(tot_i2_cvr, 0.975)`), with most variation being driven by between study and trait differences (see Supplement;  @fig-s3). Each life-history stage exhibited the same pattern of variance change in each of the habitats (@fig-cvrlh). Reduced relative variance was particularly prominant for resting metabolic rates and sprint speed although traits differed in whether they exhibited a reduction in variation in physiological rates at high temperatures (@fig-cvrtrait).

```{r, fig-cvrlh}
#| label: fig-cvrlh
#| fig-cap: "Estimated mean acclimation and acute $lnRR_{Q_{10}}$ (A-C) and $lnCVR_{Q_{10}}$ (D-F) for adult and juevenile life-history stages for Marine (A & D), Freshwater (B & E) and Terrestrial (C & F) ectotherms. k = total number of effect size estimates while the numbers in brackets indicate the number of species. Thick bars indicate 95% confidence intervals and thin bars indicate 95% prediction intervals. Raw data for both adult and jeunvile life-history stages also presented but point are not separated by type for ease of presentation."
#| echo: false
#| warning: false

@@

```{r, fig-4}
#| label: fig-4
#| fig-align: center
#| fig-cap: "Predicted mean (thick black line) $lnRR_{Q_{10}}$ as a function of the Thermal Coefficient of Variation (CV) (A) and thermal predictability (B) for wild populations across marine, freshwater and terrestrial habitats. Dashed lines indicate 95% confidence intervals and dotted lines indicate 95% prediction intervals. Model slope ($\\beta$) along with the 95% CI and $p_{MCMC}$ value are shown for each habitat."

@@

```{r, fig-5}
#| label: fig-5
#| fig-align: center
#| fig-cap: "Predicted mean (thick black line) $lnCVR_{Q_{10}}$ as a function of the Thermal Coefficient of Variation (CV) (A) and thermal predictability (B) for wild populations across marine, freshwater and terrestrial habitats. Dashed lines indicate 95% confidence intervals and dotted lines indicate 95% prediction intervals. Model slope ($\\beta$) along with the 95% CI and $p_{MCMC}$ value are shown for each habitat."

@@

Acute and acclimation responses for wild ectotherms were much less common than studies on captive populations ($N_{species}$ = `r length(unique(all_pred$spp))`, from `r length(unique(all_pred$uniques))` populations). Globally, there was a clear bias towards species in the Northern Hemisphere (@fig-fig6 A-C). Projected changes in physiological rate variance were highly variable across the globe, with some regions showing a decrease in physiological rate variance, while others showing an increase (@fig-fig6 D). However, out of the `r length(unique(all_pred2$match))` populations variance was predicted to decrease in `r (sum(decreases$n, na.rm = TRUE) / length(decreases$n[complete.cases(decreases$n)]))*100`% of the locations.

Predictions of current global changes in physiological rate variance were generally conservative with our model explaining ~ 50% of the variation in the observed data ($R^2$ = `r round(mean(r2_2), 2)`, 95% CI: `r round(quantile(r2_2, 0.025)[1], 2)` to `r round(quantile(r2_2, 0.975)[1], 2)`). Across habitat types climate change is predicted to result in a only a `r abs(lapply(mean_perc_habitat, mean)$f)`% change in variance for freshwater systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$f[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$f[2]`%, $p_{MCMC}$ = `r pmcmc(mean_perc_habitat$f)`), whereas we expect a `r abs(lapply(mean_perc_habitat, mean)$m)`% reduction in variance for marine systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$m[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$m[2]`%, $p_{MCMC}$ = `r pmcmc(mean_perc_habitat$m)`) and a `r abs(lapply(mean_perc_habitat, mean)$t)`% reduction in variance for terrestrial systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$t[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$t[2]`%, $p_{MCMC}$ = `r p_value(pmcmc(mean_perc_habitat$t))`) under a RCP8.5 climate scenerio.

@@

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011]. Here, we show that the relative variance ($lnCVR_{Q_{10}}$) in physiological rates of ectotherms is expected to decrease for ectotherms as temperatures increase, and this decrease is particularly pronouced in terrestrial ectotherms (~28%). Decreases in the relative variance of physiological rates are equal to or larger than the expected benefits from acclimation ($lnRR_{Q_{10_{acclimation}}}$) in marine and terrestrial ectotherms (*marine*: 16% acclimation vs. 15% reduction in relative variance; *terrestrial*: no acclimation vs 28% reduction in relative variance; @fig-1). Such effects are expected to result in a decrease in the variability in physiological rates for marine and terrestrial ectotherms between ~4-13% under future climate change projections. Our results provide insight into the multifaceted ways in which climate warming will impact ectotherm populations, with important ecological and evolutionary implications that require further empirical and theoretical attention. We discuss our findings in the context of the current literature and highlight key areas for future research.

### ***Potential evolutionary consequences of reduced plasticity and variance in physiologcial rates across ectotherms***

Understanding the interplay between plasticity and genetic adaptation has important implications for predicting a populations resilience to climate change [@urban2023and; @Chevin2010; @hoffmann2011climate; @merila2014climate]. Phenotypic plasticity can allow populations to persist in the face of climate extremes by matching phenotypes to fitness optima, although this is not always the case [@noble2019plastic; @radersma2020plasticity] and there may be limits and costs to plasticity that constrain its expression [@Dewitt1998; but see @Murren2015]. We show that plastic adjustments in physiological rates and changes in their relative variance differs across habitats. In freshwater habitats, acclimation responses will likely be beneficial considering that these effects are larger than reductions in relative variance in physiological rates. In contrast, acclimation responses of marine organisms coincide with a nearly equal reduction in relative variance in physiolgical rates, suggesting acclimation may not be as beneficial. Relative variance reductions are particularly pronounced in terrestrial ectotherms and this coincides with a general inability to acclimate. Considering acute responses of animals acclimated to high temperatures are generally slightly elevated compared to cold acclimated animals (~7%; $\beta$ = `r model1.s1$b[3]`, 95% CI: `r model1.s1$ci.lb[3]` to `r model1.s1$ci.ub[3]`, p `r p_value(model1.s1$pval[3])`), acclimatization is not likely to provide adaptive benefits under climate change. Together, these findings highlight the vulnerability of terrestrial ectotherms to climate change. It may be expected that reductions in variance of this magnitude impact the 'opportunity for selection' by reducing the strength of selection and/or impacting underlying genetic variance exposed to selection. Both are predicted to impact the capacity of populations to evolve under climate change [@hoffmann2011climate; @urban2023and].

@@

Theoretical models predict that plasticity should evolve in populatons experiencing greater environmental variability (spatial or temporal), particularly if oscillations are predictable over time making environmental cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogeneity in terrestrial ecosystems [@steele2019comparison] suggests that plasticity should be more likely to evolve in terrestrial environments. However, we did not find evidence to this effect. This finding is consistent with other meta-analyses [e.g., @barley2021limited; @Gunderson2015-hz] with thermal tolerance.  The lack of relationship between physiological plasticity and environmental variability and predictability could be due to the challenges in estimating realistic microthermal environments given that air temperatures may not represent thermal enviroments many taxa actually experience in nature. Alternatively, lower predictability of thermal environments over time may make cues unreliable preventing plastic responses from evolving easily [@Leung2020; @hendry2016key]. Randomly flutuating environments have been suggested to select for reduced plasticity, particularly if plastic responses are costly and this has been demonstrated by a number of empirical studies [@Leung2020; @rescan2022experimental; @leung2023phenotypic]. For example, using seed beetles (*Callosobruchus maculatus*), @hallsson2012selection showed that experimentally evolving populations under randomly fluctuating thermal conditions did not show any increase in plasticity, but had reduced plasticity. @Leung2020 also showed reduced plasticity in morphological traits of experimentally evolving algae populations (*Dunaliella salina*) when environments were less predictable [see also @leung2023phenotypic]. These findings, in combination with our results, suggests that there are costs to being plastic or that the environmental signals are insufficient to trigger endocrine and epigenetic mechanisms that lead to plasticty when environments are not predictable (either intrinsic physiological costs, fitness costs or both). The fact that aquatic ectotherms did show a capacity to acclimate suggest that environmental predictability, which is expected to be higher in aquatic environments, may be more important than environmental variability in driving the evolution of plasticity.

### ***Ecological consequences of reduced variation in physiological rates***
Variance reductions in metabolism and performance (e.g., sprint speed) were key traits most strongly impacted by increased temperature. Changes in variability in physiological rates, particularly traits governing energy demand, could have important consequences on the flow of energy within and between populations, communities and ecosystems [@hendry2016key]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation]. Maintaining intrapopulaton variability in physiological rates in a warmer world may therefore be important for population resilience to climate change. For example, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population fluctuations compared to scenarios with no variation [@kooijman1989population]. The differences in variance change across major habitat types suggest that terrestrial ectotherms will likely face greater challenges to adapting to climate change [@hoffmann2013upper; @Gunderson2015-hz; @Pottier2022], leading to greater rates of extinction and reduced productivity. However, data testing the impacts of changes in variance on populations are limited to very few taxa. We need creative experimentation across broader taxa that manipulate levels of variance to understand their ecological outcomes. Our meta-analytic findings can be integrated into theoretical models by permitting model paramaters to capture biologically relevant variation. Such predictive models can then be integrated with experimental approaches to better understand the consequences of changes in variation on ecological dynamics.

### ***Limitations***
A number of limitations need to be recognised to better contextualise our results. First, and as mentioned above, the evolutionary implications of changes in relative variances in physiological rates is still unclear because we lack a full understanding of how much variation is genetic versus environmental, or how traits are related to fitness. Second, while we found no relationships between historical climate variability / predictability and capacity for plasticity this does not necessarily mean they are not important. There are major challenges to estimating realistic microthermal environments experienced by taxa; it is possible that more fine-scale resolution of the habitats that organisms occupy will elucidate expected patterns. Third, while we show general patterns across taxa and habitats it is important to recognise that this does not mean such patterns apply to all populations. Substantial variation in acclimation responses and changes in variance exist among populations as evidenced by wide prediction intervals. As such, finer scale analyses that look at acclimation responses *and* changes in relative variance will be beneficial for developing a better understanding of their interplay when considering how populations will be impacted by climate change.

@@

Enhanced knowledge of how variation in physiological rates vary across populations and species and the degree to which they can be adjusted in response to the environment may increase our ability to make more informed predictions about the ecological and evolutionary dynamics of natural populations [@forsman2015rethinking]. Changes in the relative variance in physiological rates could be better incorporated into physiological and ecological models to provide more nuanced and possibly more realistic predictions about the impacts of climate change on natural populations. While we do not yet understand the relative contribution of environmental and genetic factors to variances changes, models could better decouple how different levels of heritability with different total variance impact evolutionary and ecological predictions. Our meta-analysis allows for realistic models to be parameterised and ensure they are better aligned with empirical findings. Nonetheless, many fascinating questions remain unanswered that will require greater focus on the consequences of changes in variance (rather than just the mean). Particularly interesting questions include: How does a reduction in physiological rate variance change energy flow across tropic levels within communities? What are the biochemical, cellular and physiological mechanisms that underly reduced variance in physiological rates? Are reductions in variance in one trait associated with reductions in correlated traits, or do some traits increase while others decrease? How much of the reduction in variance is driven by lower levels of genetic variance? Answers to these questions will require integrative approaches that combine empirical and theoretical work across multiple levels of biological organisation but will likely provide useful advances in understanding the full consequences that climate change will have on ectotherms across all major ecosystems.

>>>>

Climate change is expected to result in warmer and more variable thermal environments globally. Greater thermal variability is expected to result in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity. Such responses depend on genetic and phenotypic variability, yet most work has focused on plastic changes in mean phenotypic responses to climate warming ignore how temperature may change phenotypic variability. Here, using a large database of physiological rate data from marine, freshwater and terrestrial ectotherms we quantify the capacity for temperature acclimation and the extent to which relative variance (changes in variance after accounting for changes in the mean, or coefficient of variation) in physiological rates changes in response to increasing temperatures. Using new effect size estimates and meta-analysis we show that acclimation of physiological rates and changes in their relative variance do not depend on past temperature variability and predictability experienced by a population at its geographical location; neither do these responses differ between life stages. Across habitats relative variance in physiological rates decreased at higher temperatures. Freshwater ectotherms are capable of acclimating which is likely to be beneficial considering that reductions in relative variance in physiological rates at high temperatures is lower for freshwater species. In contrast, acclimation responses of marine organisms coincide with a nearly equal reduction in relative variance in physiological rates, suggesting acclimation may not necessarily be beneficial. Relative variance reductions are particularly pronounced for terrestrial ectotherms and this coincides with a lack of capacity for acclimation, which highlights the vulnerability of terrestrial ectotherms to climate change. Our results emphasise how beneficial acclimation responses may be offset by reductions in physiological rate variance, challenging the notion that plasticity will be sufficient to counteract climate change effects on populations. Reductions in physiological rate variance at high temperatures may further have important ecological ramifications that impact population and ecosystem processes that cascade to affect population and community productivity.

@@

Reversible phenotypic plasticity, such as acclimation, is expected to provide greater potential to buffer populations from climate impacts as responses are relatively rapid and can therefore be fine-tuned to proximate environmental conditions [assuming the costs of plasticity are low – @Dewitt1998; @Scheiner2020]. Acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to facilitate a rapid response to the environment [@seebacher2019histone; @little2013thyroid; @taff2016endocrine]. However, the focus up to now has been primarily on mean physiological responses. For example, mean thermal tolerances or acclimation abilities in a population are likely to shift in response to thermal environments [@Seebacher2015-wi; @Gunderson2015-hz; @Havird2020-mj; @Pottier2022]. However, it is possible that intrapopulation variability might also be impacted in addition to the mean. Understanding how variability in physiological rates -- traits thought to be closely linked to fitness -- are affected by climate change is important because lack of physiological variation can limit responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variance in a population may also indicate greater niche breadth which can buffer populations against environmental change [@Bolnick2011; @Schindler2010; @Zheng2023; @Slatyer2013]. Decreases in phenotypic variance also suggests strong stabilising selection or reflect constraints on performance [@Scheiner2020; @Dewitt1998]. Changes in physiological trait variation may also have important ecological consequences by promoting population productivity and stability [@Agashe2009; @kooijman1989population], species coexistence and ecosystem processes [@imura2003genetic; @Bolnick2011; @hart2016variation]. The implications of changes in variance could, therefore, have wide-reaching consequences for understanding the capacity of populations to persist in and adapt to novel environments but to date have been largely ignored [@Scheiner2020].

@@

Nonetheless, effect heterogeneity was high (only `r mean(tot_i2)`% of the variance was the result of sampling variability, 95% CI: `r quantile(tot_i2, 0.025)[1]` to `r quantile(tot_i2, 0.975)[1]`%), and most variance was explained by the specific study and type of trait (Study: `r mean(stdy_i2)`% , 95% CI: `r quantile(stdy_i2, 0.025)[1]` to `r quantile(stdy_i2, 0.975)[1]`%; Trait Type: `r mean(trait_i2)`% , 95% CI: `r quantile(trait_i2, 0.025)[1]` to `r quantile(trait_i2, 0.975)[1]`%). Evolutionary relationships among taxa and species ecology (i.e., species random effect) explained little variation in acute and acclimation responses (Species: `r mean(spp_i2)`% , 95% CI: `r quantile(spp_i2, 0.025)[1]` to `r quantile(spp_i2, 0.975)[1]`%; Phylogeny: `r mean(phylo_i2)`% , 95% CI: `r quantile(phylo_i2, 0.025)[1]` to `r quantile(phylo_i2, 0.975)[1]`%).

@@

```{r, fig-1}
#| label: fig-1
#| fig-cap: "Taxonomic distribution of acute and acclimation $Q_{10}$ estimates across major habitats. **A)** Phylogenetic distribution of taxa contained within the data. The total number of acute and acclimation type Q10 effect sizes are highlighted as well as whether the taxa is marine, freshwater or terrestrial. Silhouettes are representative taxa of major clades within the tree. **B)** Mean acute and acclimation $lnRR_{Q_{10}}$ across marine, freshwater, and terrestrial systems. **C)** Mean $lnCVR_{Q_{10}}$ across traits for marine, freshwater and terrestrial systems. Note there were no differences between acute and acclimation $Q_{10}$ types. k = total number of effect size estimates while the numbers in brackets indicate the number of species. Thick bars are 95% confidence intervals (CI) and thin bars 95% prediction intervals (PI)."
#| warning: false
#| echo: false

@@

The total proportion of heterogeneity in $lnCVR_{Q_{10}}$ was lower compared to $lnRR_{Q_{10}}$ ($I^2_{sv}$ = `r mean(tot_i2_cvr)`, 95% CI: `r quantile(tot_i2_cvr, 0.025)` to `r quantile(tot_i2_cvr, 0.975)`), with most variation being driven by between study and trait differences (see Supplement;  @fig-s3). Each life-history stage exhibited the same pattern of variance change in each of the habitats (@fig-cvrlh). Reduced relative variance was particularly prominent for resting metabolic rates and sprint speed although traits differed in whether they exhibited a reduction in variation in physiological rates at high temperatures (@fig-cvrtrait).

```{r, fig-cvrlh}
#| label: fig-cvrlh
#| fig-cap: "Estimated mean acclimation and acute $lnRR_{Q_{10}}$ (A-C) and $lnCVR_{Q_{10}}$ (D-F) for adult and juvenile life-history stages for Marine (A & D), Freshwater (B & E) and Terrestrial (C & F) ectotherms. k = total number of effect size estimates while the numbers in brackets indicate the number of species. Thick bars indicate 95% confidence intervals and thin bars indicate 95% prediction intervals. Raw data for both adult and juvenile life-history stages also presented but point are not separated by type for ease of presentation."
#| echo: false
#| warning: false

@@

```{r, fig-4}
#| label: fig-4
#| fig-align: center
#| fig-cap: "Predicted mean (thick black line) $lnRR_{Q_{10}}$ as a function of the Thermal Coefficient of Variation (CV) (A) and thermal predictability (B) for wild populations across marine, freshwater and terrestrial habitats. Dashed lines indicate 95% confidence intervals and dotted lines indicate 95% prediction intervals. Model slope ($\\beta$) along with the 95% CI and $p_{MCMC}$ value is shown for each habitat."

@@

```{r, fig-5}
#| label: fig-5
#| fig-align: center
#| fig-cap: "Predicted mean (thick black line) $lnCVR_{Q_{10}}$ as a function of the Thermal Coefficient of Variation (CV) (A) and thermal predictability (B) for wild populations across marine, freshwater and terrestrial habitats. Dashed lines indicate 95% confidence intervals and dotted lines indicate 95% prediction intervals. Model slope ($\\beta$) along with the 95% CI and $p_{MCMC}$ value is shown for each habitat."

@@

Acute and acclimation responses for wild ectotherms were much less common than studies on captive populations ($N_{species}$ = `r length(unique(all_pred$spp))`, from `r length(unique(all_pred$uniques))` populations). Globally, there was a clear bias towards species in the Northern Hemisphere (@fig-fig6 A-C). Projected changes in physiological rate variance were highly variable across the globe, with some regions showing a decrease in physiological rate variance, while others showing an increase (@fig-fig6 D). However, out of the `r length(unique(all_pred2$match))` populations, relative variance was predicted to decrease in `r (sum(decreases$n, na.rm = TRUE) / length(decreases$n[complete.cases(decreases$n)]))*100`% of the locations.

Predictions of current global changes in physiological rate variance were generally conservative with our model explaining ~ 50% of the variation in the observed data ($R^2$ = `r round(mean(r2_2), 2)`, 95% CI: `r round(quantile(r2_2, 0.025)[1], 2)` to `r round(quantile(r2_2, 0.975)[1], 2)`). Across habitat types climate change is predicted to result in a only a `r abs(lapply(mean_perc_habitat, mean)$f)`% change in variance for freshwater systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$f[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$f[2]`%, $p_{MCMC}$ = `r pmcmc(mean_perc_habitat$f)`), whereas we expect a `r abs(lapply(mean_perc_habitat, mean)$m)`% reduction in relative variance for marine systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$m[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$m[2]`%, $p_{MCMC}$ = `r pmcmc(mean_perc_habitat$m)`) and a `r abs(lapply(mean_perc_habitat, mean)$t)`% reduction in relative variance for terrestrial systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$t[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$t[2]`%, $p_{MCMC}$ = `r p_value(pmcmc(mean_perc_habitat$t))`) under a RCP8.5 climate scenerio.

@@

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011]. Here, we show that the relative variance ($lnCVR_{Q_{10}}$) in physiological rates for ectotherms is expected to decrease as temperatures increase, and this decrease is particularly pronouced in terrestrial ectotherms (~28%). Decreases in the relative variance of physiological rates are equal to or larger than the expected benefits from acclimation ($lnRR_{Q_{10_{acclimation}}}$) in marine and terrestrial ectotherms (*marine*: 16% acclimation vs. 15% reduction in relative variance; *terrestrial*: no acclimation vs 28% reduction in relative variance; @fig-1). Such effects are expected to result in a decrease in the variability in physiological rates for marine and terrestrial ectotherms between ~4-13% under future climate change projections. Our results provide insight into the multifaceted ways in which climate warming will impact ectotherm populations, with important ecological and evolutionary implications that require further empirical and theoretical attention. We discuss our findings in the context of the current literature and highlight key areas for future research.

### ***Potential evolutionary consequences of reduced plasticity and variance in physiological rates across ectotherms***

Understanding the interplay between plasticity and genetic adaptation has important implications for predicting a populations resilience to climate change [@urban2023and; @Chevin2010; @hoffmann2011climate; @merila2014climate]. Phenotypic plasticity can allow populations to persist in the face of climate extremes by matching phenotypes to fitness optima, although this is not always the case [@noble2019plastic; @radersma2020plasticity] and there may be limits and costs to plasticity that constrain its expression [@Dewitt1998; but see @Murren2015]. We show that plastic adjustments in physiological rates and changes in their relative variance differs across habitats. In freshwater habitats, acclimation responses will likely be beneficial considering that these effects are larger than reductions in relative variance in physiological rates. In contrast, acclimation responses of marine organisms coincide with a nearly equal reduction in relative variance in physiological rates, suggesting acclimation may not be as beneficial. Relative variance reductions are particularly pronounced in terrestrial ectotherms, and this coincides with a general inability to acclimate. Considering acute responses of animals acclimated to high temperatures are generally slightly elevated compared to cold-acclimated animals (~7%; $\beta$ = `r model1.s1$b[3]`, 95% CI: `r model1.s1$ci.lb[3]` to `r model1.s1$ci.ub[3]`, p `r p_value(model1.s1$pval[3])`), acclimatization is not likely to provide adaptive benefits under climate change. Together, these findings highlight the vulnerability of terrestrial ectotherms to climate change. Theoretical models that capture the percentage changes in relative variances and acclimation responses we observe here are needed fully appreciate the biological consequences such changes have on ecological and evolutionary dynamics. It may be expected that reductions in variance of this magnitude impact the 'opportunity for selection' by reducing the strength of selection and/or impacting underlying genetic variance exposed to selection. Both are predicted to impact the capacity of populations to evolve under climate change [@hoffmann2011climate; @urban2023and].

@@

Theoretical models predict that plasticity should evolve in populations experiencing greater environmental variability (spatial or temporal), particularly if oscillations are predictable over time, making environmental cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogeneity in terrestrial ecosystems [@steele2019comparison] suggests that plasticity should be more likely to evolve in terrestrial environments. However, we did not find evidence to this effect. This finding is consistent with other meta-analyses [e.g., @barley2021limited; @Gunderson2015-hz] with thermal tolerance.  The lack of relationship between physiological plasticity and environmental variability and predictability could be due to the challenges in estimating realistic microthermal environments given that air temperatures may not represent thermal enviroments many taxa actually experience in nature. Therefore, it is possible that more fine-scale resolution of the habitats that organisms occupy will elucidate expected patterns. Alternatively, lower predictability of thermal environments over time may make cues unreliable, preventing plastic responses from evolving easily [@Leung2020; @hendry2016key]. Randomly fluctuating environments have been suggested to select for reduced plasticity, particularly if plastic responses are costly and this has been demonstrated by a number of empirical studies [@Leung2020; @rescan2022experimental; @leung2023phenotypic]. For example, using seed beetles (*Callosobruchus maculatus*), @hallsson2012selection showed that experimentally evolving populations under randomly fluctuating thermal conditions did not show any increase in plasticity, but had reduced plasticity. @Leung2020 also showed reduced plasticity in morphological traits of experimentally evolving algae populations (*Dunaliella salina*) when environments were less predictable [see also @leung2023phenotypic]. These findings, in combination with our results, suggest that there are costs to being plastic or that the environmental signals are insufficient to trigger endocrine and epigenetic mechanisms that lead to plasticity when environments are not predictable. The fact that aquatic ectotherms did show a capacity to acclimate suggests that environmental predictability, which is expected to be higher in aquatic environments, may be more important than environmental variability in driving the evolution of plasticity.

### ***Ecological consequences of reduced variation in physiological rates***
Variance reductions in metabolism and performance (e.g., sprint speed) were key traits most strongly impacted by increased temperature. Changes in variability in physiological rates, particularly traits governing energy demand, could have important consequences on the flow of energy within and between populations, communities, and ecosystems [@hendry2016key]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation]. Maintaining intrapopulation variability in physiological rates in a warmer world may therefore be important for population resilience to climate change. For example, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population fluctuations compared to scenarios with no variation [@kooijman1989population]. The differences in variance change across major habitat types suggest that terrestrial ectotherms will likely face greater challenges to adapting to climate change [@hoffmann2013upper; @Gunderson2015-hz; @Pottier2022], leading to greater rates of extinction and reduced productivity. However, data testing the impacts of changes in variance on populations are limited to very few taxa. We need creative experimentation across broader taxa that manipulate levels of variance to understand their ecological outcomes. Our meta-analytic findings can be integrated into theoretical models by permitting model parameters to capture biologically relevant variation. Such predictive models can then be integrated with experimental approaches to better understand the consequences of changes in variation on ecological dynamics.

@@

Enhanced knowledge of how variation in physiological rates vary across populations and species and the degree to which they can be adjusted in response to the environment may increase our ability to make more informed predictions about the ecological and evolutionary dynamics of natural populations [@forsman2015rethinking]. While we show general patterns across taxa and habitats it is important to recognise that this does not mean such patterns apply to all populations. Substantial variation in acclimation responses and changes in variance exist among populations as evidenced by wide prediction intervals. Nonetheless, changes in the relative variance in physiological rates could be better incorporated into physiological and ecological models to provide more nuanced and possibly more realistic predictions about the impacts of climate change on natural populations. While we do not yet understand the relative contribution of environmental and genetic factors to variances changes, models could better decouple how different levels of heritability with different total variance impact evolutionary and ecological predictions. Our meta-analysis allows for realistic models to be parameterised and ensure they are better aligned with empirical findings. Nonetheless, many fascinating questions remain unanswered that will require greater focus on the consequences of changes in variance (rather than just the mean). Particularly interesting questions include: How does a reduction in physiological rate variance change energy flow across tropic levels within communities? What are the biochemical, cellular and physiological mechanisms that underly reduced variance in physiological rates? Are reductions in variance in one trait associated with reductions in correlated traits, or do some traits increase while others decrease? How much of the reduction in variance is driven by lower levels of genetic variance? Answers to these questions will require integrative approaches that combine empirical and theoretical work across multiple levels of biological organisation but will likely provide useful advances in understanding the full consequences that climate change will have on ectotherms across all major ecosystems.
